---
title: "Stanford Open Policing EDA"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(stringr)
d = read_csv('WI-clean.csv', col_types = list(county_fips = 'c', officer_id = 'c'))
```

The [README](https://github.com/5harad/openpolicing/blob/master/DATA-README.md) describes the data in more detail. The datasets provided are cleaned versions in a fairly tidy format; each row denotes one stop, each column denotes a variable, and each cell denotes a value. Columns denoted with \*_raw are original data values, which are standardized to the corresponding \* columns. Since there are differences across datasets from state to state, the WI dataset is explored more in depth.

## A first glimpse

The dataset contains 1,059,033 rows and 27 columns.
```{r}
d %>% glimpse
```
```{r}
d %>% summary()
```

## id is the primary key
The variable id is the primary key which uniquely identifies each stop.
```{r}
d %>% 
  count(id) %>% 
  filter(n > 1)
```

## state all WI
Each observation's state is WI. The [documentation](https://github.com/5harad/openpolicing/blob/master/DATA-README.md) states this variable is "the two-letter code for the state in which the stop occurred." Note that this conflicts with some results found by looking at latitude/longitude, where some stops were outside WI borders in neighboring states.
```{r}
# each stop was in WI
d %>% 
  group_by(state) %>%
  count()
nrow(d)
```

## Stop dates
The following explores the number of stops made by day, year, and month.
```{r}
# days in dataset with most stops
d %>%
  group_by(stop_date) %>% 
  count() %>% 
  arrange(desc(n))
```

The amount of stops for 2011 - 2015 is comparable. There are less stops in the dataset for 2010, and the data for 2016 stops at 5/16/2016.
```{r}
# most stops by year
d %>%
  select(stop_date) %>% 
  transmute(yr = year(stop_date)) %>% 
  group_by(yr) %>% 
  count() %>% 
  arrange(desc(n))
```

Months with largest amount of stops. Note this includes 2010 and 2016.
```{r}
# most stops by month
d %>%
  select(stop_date) %>% 
  transmute(month = month(stop_date)) %>% 
  group_by(month) %>% 
  count() %>% 
  arrange(desc(n))
```

Days of the week with largest amount of stops. Note this includes 2010 and 2016.
```{r}
# most stops by day of the week
d %>%
  select(stop_date) %>% 
  transmute(day = wday(stop_date, label = TRUE)) %>% 
  group_by(day) %>% 
  count() %>% 
  arrange(desc(n))
```

Days of the month with largest amount of stops. Note this includes 2010 and 2016.
```{r}
# most stops by day of the month
d %>%
  select(stop_date) %>% 
  transmute(day = mday(stop_date)) %>% 
  group_by(day) %>% 
  count() %>% 
  arrange(desc(n))
```

## Stop times
Which hours are the most number of stops made?
```{r}
d %>% 
  group_by(hour(stop_time)) %>% 
  count() %>% 
  arrange(desc(n))
```

## Geographic: location_raw, county_name, county_fips, fine_grained_location
The counties with the highest frequency of stops. The column county_name is the name of the county, while county_fips is the corresponding FIPS code.
```{r}
d %>%
  group_by(county_name) %>% 
  count() %>% 
  arrange(desc(n))
d %>%
  group_by(county_fips) %>% 
  count() %>% 
  arrange(desc(n))
```

Note county_name = location_raw + " County". These columns are repetitive, so one of these can be dropped. location_raw is the original values from which the standardized county name and FIPS code is obtained.
```{r}
# location_raw is part of county_name before " County"
d %>%
  select(county_name) %>% 
  separate(county_name, into = c("county", "word_county"), sep = -8) %>%
  select(county) %>% 
  mutate(county = toupper(county)) %>% 
  identical(d %>% select(location_raw) %>% transmute(county = location_raw))
```
```{r}
# the last word of county_name is always "County"
d %>%
  select(county_name) %>% 
  separate(county_name, into = c("county", "word_county"), sep = -8) %>% 
  group_by(word_county) %>% 
  count()
```

More specific information about the location is provided with fine_grained_location. Below is a simplistic view of where the largest number of stops occurred. Note there are many NAs and more in-depth analysis can be done.
```{r}
# the last word of county_name is always "County"
d %>%
  group_by(fine_grained_location) %>% 
  count() %>% 
  arrange(desc(n))
```

## Police department
Departments which made the most number of stops. We speculate that Wisconsin State Patrol is a general category, while WI State Patrol <region>/<office> are more specific. The [Wisconsin DOT website](http://wisconsindot.gov/Pages/about-wisdot/who-we-are/dsp/loc-contact.aspx) indicates the various regions -- Southwest, Southeast, Northeast, North Central, Northwest, which correspond to the abbreviations SWR, SER, NER, NCR, and NWR below, respectively. Based on the same page previously and the [Wikipedia page](https://en.wikipedia.org/wiki/Wisconsin_State_Patrol#The_State_Patrol_today), there are "offices in DeForest (Madison), Waukesha, Fond du Lac, Wausau, Tomah, Eau Claire and Spooner", which correspond to the abbreviations DEF, WKE, FON, WSA, TOM, EAU, and SPO.

It is unclear what ZWI State Patrol refers to; this may be some sort of error in the data.
```{r}
d %>% 
  group_by(police_department) %>% 
  count() %>% 
  arrange(desc(n))
```

## Driver gender
Males in WI were stopped more often than females. There are also a significant number of NAs.
```{r}
d %>% 
  group_by(driver_gender) %>% 
  count() %>% 
  arrange(desc(n))
```


## Driver age
Driver age information is all NA.
```{r}
d %>% 
  group_by(driver_age_raw) %>% 
  count()
d %>% 
  group_by(driver_age) %>% 
  count()
```

## Driver race
```{r}
# driver_race_raw -> driver_race
# A -> Asian
# B -> Black
# H -> Hispanic
# I -> Other
# W -> White
# O + NA -> NA
# What is O? 
d %>%
  group_by(driver_race_raw) %>% 
  count() %>% 
  arrange(desc(n))
d %>%
  group_by(driver_race) %>% 
  count() %>% 
  arrange(desc(n))
```

## Violations
violation_raw denotes the original data values, while violation is the standardized category of the violation. Note that some stops contain multiple violations; to explore this more in-depth, cases with multiple violations could be split.
```{r}
# most common violations (specific)
d %>% 
  group_by(violation_raw) %>% 
  count() %>% 
  arrange(desc(n))
# most common violations (general)
d %>% 
  group_by(violation) %>% 
  count() %>% 
  arrange(desc(n))
```

## Search type
```{r}
# most common search types
d %>% 
  group_by(search_type) %>% 
  count() %>%
  arrange(desc(n))
```
Note that search_type_raw and search_type are the same.
```{r}
identical(d$search_type, d$search_type_raw)
```

## Stop outcome
For stops with multiple outcomes, the value is the most severe outcome.
```{r}
d %>% 
  group_by(stop_outcome) %>% 
  count() %>%
  arrange(desc(n))
```

## Latitude, longitude
According to the Wikipedia page for Wisconsin, latitude ranges from 42°30' N to 47°05′ N, while longitude ranges from 86°46′ W to 92°54′ W. There appear to be some errors in values of latitude and longitude, with some values not fitting in the range suggested on Wikipedia either in magnitude and/or direction (some values of longitude are positive and therefore E, not W).

Filtering the latitude by this range reveals that there appears to be some errors as shown below, with some latitude values being too small or large. For instance, the latitude 440721.0000 may have been intended to be 44.0721. (The row with latitude 440721.0000	and longitude 874232.00000 may be intended to be latitude 44.0721 and longitude -87.4232, which maps to a position in Lake Michigan near a line Google Maps labels as Ludington Manitowoc. This seems to be some kind of bridge/ferry?)
```{r}
d %>% 
  filter(lat < 42.3 | lat > 47.05) %>% 
  arrange(lat) %>% 
  select(id, lat, lon, everything())
```

There are some longitude values which are positive, corresponding to E instead of W. These should most likely be negative instead of positive; otherwise this maps to locations on the other side of the world in areas like China.
```{r}
d %>% 
  filter(lon < -92.54 | lon > -86.46) %>% 
  arrange(lat) %>% 
  select(id, lat, lon, everything())
```

There are also locations which map to locations outside WI borders into neighboring states. For instance, there are many stops with the latitude 44.91696 and longitude 92.86195. However, this corresponds to a location in northern China, near the border with Mongolia. This seems to be an error, with the intended longitude of -92.86195. The latitude 44.91696 and longitude -92.86195 maps to a location in eastern Minnesota, which is not too far from the western border to Wisconsin. Further inspection reveals that officer with id 1787 seems to frequent this area, as the officer has made 469 stops within the time span of the dataset.
```{r}
d %>% 
  filter(lat == 44.91696 | lon == 92.86195) %>% 
  group_by(officer_id) %>% 
  count()
```

## Officer
```{r}
d %>% 
  group_by(officer_id) %>% 
  count() %>% 
  arrange(desc(n))
```

## Vehicle type
This column seems to be in the format (manufacturer, model, year). The following is a simple first look at what combination of these characteristics is most frequent. To explore this in more depth, this could be separated into separate columns for manufacturer, model, and year.
```{r}
d %>% 
  group_by(vehicle_type) %>% 
  count() %>% 
  arrange(desc(n))
```



## Missing race data by officer
The following is sorted in descending order by number of stops with missing race an officer has made, and missingRaceRatio indicates what proportion of their total stops is missing race.
```{r}
d %>% 
  mutate(missingRace = is.na(driver_race)) %>% 
  group_by(officer_id) %>% 
  summarise(missingRaceRatio = mean(missingRace == TRUE), numMissing = sum(missingRace), numStops = n()) %>% 
  arrange(desc(numMissing))
```

## Stops with missing race data tend to miss other information about stop
It seems to be the case that stops with missing information about driver race often are missing other information about the stop overall, including driver gender, whether or not a search was conducted, stop time, and information about the vehicle (manufacturer, model, and year all missing). Unfortunately, all values for driver age are missing in the dataset, otherwise that could be explored as well. 
```{r}
missingRace <- d %>% filter(is.na(driver_race))
sprintf("Number of stops with missing race: %s", missingRace %>% nrow()) 
sprintf("Number of stops with missing race, gender, search conducted, stop time, vehicle information: %s", d %>% filter(is.na(driver_race), is.na(driver_gender), is.na(search_conducted), is.na(stop_time), vehicle_type == "NA NA NA") %>% nrow())
```
For instance, stops missing driver race is often missing driver gender.
```{r}
d %>% 
  mutate(missingRace = is.na(driver_race)) %>%
  group_by(driver_gender) %>% 
  summarise(missingRaceRatio = mean(missingRace == TRUE), numMissing = sum(missingRace), numStops = n()) %>% 
  arrange(desc(numMissing))
```
Stops missing whether a search was conducted is often missing driver race.
```{r}
d %>% 
  mutate(missingRace = is.na(driver_race)) %>%
  group_by(search_conducted) %>% 
  summarise(missingRaceRatio = mean(missingRace == TRUE), numMissing = sum(missingRace), numStops = n()) %>% 
  arrange(desc(numMissing))
```
  
## Missing race by year: Proportion of stops with missing race data increases over time
Note that 2010 is thrown out since there is less data about stops in 2010 (< 6000). It seems that the proportion of stops with missing race information is increasing as time goes on; earlier years like 2011 and 2012 have more information regarding race, while later years like 2015 and 2016 have more missing race more frequently. Most years have a comparable amount of stops (all around 200,000, except for 2016 since it is missing the last several months).
```{r}
missingRaceByYear <- d %>% 
  mutate(missingRace = is.na(driver_race)) %>% 
  mutate(yr = year(stop_date)) %>% 
  filter(!is.na(yr), yr != 2010) %>% 
  group_by(yr) %>% 
  summarise(missingRaceRatio = mean(missingRace == TRUE), numMissing = sum(missingRace), numStops = n()) %>% 
  arrange(desc(missingRaceRatio))
missingRaceByYear
missingRaceByYear %>% ggplot(mapping = aes(x = yr, y = missingRaceRatio)) +
    geom_line()
```

## Missing race by county
Race information seems to be missing fairly uniformly throughout counties. That is, the number of missing stops roughly follows the pattern of counties with a large number of stops overall.
```{r}
d %>% 
  mutate(missingRace = is.na(driver_race)) %>%
  group_by(county_name) %>% 
  summarise(missingRaceRatio = mean(missingRace == TRUE), numMissing = sum(missingRace), numStops = n()) %>% 
  arrange(desc(numMissing))
```

## Missing race data by police department
Stops where the police_department is Wisconsin State Patrol have a surprisingly small amount of missing race values, while stops where the police_department is more specific tend to have higher instances of race missing. Police departments SWR/TOM and NWR/SPO has higher proportion of missing race data, while the other departments have around half of their stops missing race information.
```{r}
d %>% 
  mutate(missingRace = is.na(driver_race)) %>%
  group_by(police_department) %>% 
  summarise(missingRaceRatio = mean(missingRace == TRUE), numMissing = sum(missingRace), numStops = n()) %>% 
  arrange(desc(numMissing))
```
  

# getting manufacturer of vehicle 
```{r}
x <- d %>% 
  separate(vehicle_type, into = c("manu", "model", "year"), sep = " ") %>% 
  select(manu, model, year) %>% 
  group_by(manu) %>% 
  count() %>%
  arrange(desc(n))
```
  
  
# getting data in format for model fitting
# filters out 2010 data
# filters out 6 rows with NA for violation
# assume num of violations = 1 + (number of commas) = 1 + <# of ",">
# no rows are missing county_fips, so an inner join is safe (doesn't drop any rows)
# keeps relevant columns
```{r}
data <- d %>% 
  mutate(missingRace = is.na(driver_race)) %>% 
  select(stop_date, county_fips, police_department, driver_gender, violation, search_conducted, stop_outcome, missingRace) %>%
  mutate(year = year(stop_date), month = month(stop_date)) %>%
  filter(year != 2010 | is.na(year)) %>% 
  filter(!is.na(violation)) %>% 
  mutate(num_violations = 1 + str_count(violation, ",")) %>% 
  select(-stop_date, -violation) %>% 
  inner_join(extraData %>% mutate(county_fips = as.character(county_fips)), by = "county_fips") %>% 
  replace(is.na(.), "missing") %>% 
  mutate_if(is.character,as.factor) %>% 
  mutate_if(is.logical, as.factor) %>% 
  mutate(year = as.factor(year), month = as.factor(month))
```

```{r}
model <- glm(missingRace ~ driver_gender + county_fips, family = binomial, 
    data = data)
summary(model)
```

```{r}
model2 <- glm(missingRace ~ driver_gender + police_department + year, data = data, family = binomial)
summary(model2)
```